\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-15 metatheory.tex

Though phrased in Latex, we are following \citep{kaposi2017type} to use as our meta-logic a type theoretical framework (like Agda) , 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Each
judgement will actually be represented by a QIIT type, which are
mutually recursively defined together. Quotient is used to represent
judgemental(definitional) equality, thanks to which we can have a
concise representation. This way, for all four kinds of judgement we are
using, there are always equality among these data. Doing so, we can omit
some obvious equality judgement. Most of the reasoning can be formulated
using the \textit{algebra} of QIIT (mapping out function from this
quotient data-type).

>>>>

Though phrased in Latex, we are following \citet{altkap2016} to use as
our meta-logic a type theoretical framework (like Agda), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Each
judgement will actually be represented by a QIIT type, which are
mutually recursively defined together. Quotient is used to represent
judgemental(definitional) equality, thanks to which we can have a
concise representation. This way, for all four kinds of judgement we are
using, there are always equality among these data. Doing so, we can omit
some obvious equality judgement. Most of the reasoning can be formulated
using the \textit{algebra} of QIIT (mapping out function from this
quotient data-type).

### 2022-06-15 syntax.tex

\begin{figure}[H]
  \begin{minipage}[b]{0.3\linewidth}
      $$
      \Rule[name=Tm]
      {\goodCtx{\Gamma}{i} \quad \goodType{\Gamma}{T}{j}}
      {\goodTerm{\Gamma}{\_}{T}}
      $$
      $$
      \Rule[name=Tms]
      {\goodCtx{\Gamma}{i} \quad \goodCtx{\Delta}{i}}
      {\goodSub{\Gamma}{\_}{\Delta} \text{ is well-formed}}
      $$
      $$
      \Rule[name=Sig]
      {\goodCtx{\Gamma}{i} \quad n \in \nat}
      {\goodSig{\Gamma}{\_}{n} \text{ is well-formed}}
      $$
      $$
      \Rule[name=Seal]
      {\goodCtx{\Gamma}{i} \quad \goodSig{\Gamma}{\sigma}{n} 
      \quad \goodType{\Gamma}{A}{} }
      {\goodSeal{\Gamma}{\_}{\sigma}{A} \text{ is well-formed}}
      $$
  \end{minipage}
  \begin{minipage}[b]{0.6\linewidth}
    \begin{minted}[]{agda}
      data Con   : Set 
      data Ty    : Con → Set   
      data Tms   : Con → Con → Set 
      data Tm    : (Γ : Con) → Ty Γ  → Set 
      data Sig  : Con → ℕ → Set
      data WSig : Con → ℕ → Set 
      data Inh  : (Γ : Con) → Sig Γ n → Sig Γ m → Set
      Seal : (Γ : Con) → Sig Γ n → Sig Γ n → Set
    \end{minted}
  \end{minipage}
\begin{align*}
  &\Gamma~:~"Con"&&\mathboxtext{\goodCtx{\Gamma}{}} &&T~:~"Ty"~\Gamma&&\mathboxtext{\goodType{\Gamma}{T}{}}  &&\sigma~:~"Tms"~\Gamma~\Delta&&\mathboxtext{\goodSub{\Gamma}{\sigma}{\Delta}}  \\ & t~:~"Tm"~\Gamma~T&&\mathboxtext{\goodTerm{\Gamma}{t}{T}} 
  &&\sigma~:~"Sig"~\Gamma~n&&\mathboxtext{\goodSig{\Gamma}{\sigma}{n}}  &&\sigma~:~"WSig"~\Gamma~n&&\mathboxtext{\goodWSig{\Gamma}{\sigma}{n}} \\ &h~:~"Inh"~\Gamma~\sigma~\tau&&\mathboxtext{\goodInh{\Gamma}{h}{\sigma}{\tau}} && f~:~"Seal"~\Gamma~\sigma~A&&\mathboxtext{\goodSeal{\Gamma}{f}{\sigma}{A}}
\end{align*}
\caption{Well-formed-ness of Four New Judgements, and Agda-encoding-Judgements correspondence}
\end{figure}

@@

Now we show case some of the typing rules, we will have dependent function type and dependent pair type. Dependent pair type is used to model Coq module, which is useful when we need to model the ``compilation'' from linkages to modules. We also have $\beta,\eta$-rule for both dependent function type and dependent pair type.

@@

$$
\Rule[name=Ind Type]
{\goodTerm{\Gamma}{T}{\bW \sigma}}
{\goodTerm{\Gamma}{"W" \ T}{\cU}}
\quad
\Rule[name=Ind Term]
{\goodTerm{\Gamma}{T}{\bW \sigma}
  \quad \goodTerm{\Gamma}{a}{\pi^j_1\sigma}
  \quad \goodTerm{\Gamma, \pi^j_1\sigma}{b}{\pi^j_2\sigma}}
{\goodTerm{\Gamma}{"Wsup" \ T \ a \ b}{El\ ("W" \ T)} }
$$

@@

In our meta-theory, we only formulate non-dependent eliminator, i.e. the \textit{recursor}. Each recursor is constructed by a linkage contained with functions/handlers/pattern-matching cases dealing with each constructor. Thus code-reuse for recursor is delegated by the inheritance of linkage -- to reuse one particular pattern-matching case, we just inherit the function/handler. This greatly simplifies the meta\-theory development.

@@

As we said earlier, $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is actually just a short hand of term judgement $\goodTerm{\Gamma , \cC \sigma}{\_}{A[\pi_1]}$. The substitution law should understand as a lemma, and easily derivable by using $\gamma^\uparrow$.

@@

With the definition of sealing judgement in mind, we can see \ruleref{Sig Add} and \ruleref{Lnkg Add} work exactly as the implementation. In the implementation, we will have "self__" all around when defining new fields in our plugin, and this "self__" is exactly the $A$ in the context of the two fields -- especially when we don't do any sealing, this $A = \cC \sigma$, which is exactly how the implementation expose "self__" as a module parameter.

>>>>

\begin{figure}[H]
  \begin{minipage}[b]{0.3\linewidth}
      $$
      \Rule[name=Tm]
      {\goodCtx{\Gamma}{i} \quad \goodType{\Gamma}{T}{j}}
      {\goodTerm{\Gamma}{\_}{T}}
      $$
      $$
      \Rule[name=Tms]
      {\goodCtx{\Gamma}{i} \quad \goodCtx{\Delta}{i}}
      {\goodSub{\Gamma}{\_}{\Delta} \text{ is well-formed}}
      $$
      $$
      \Rule[name=Sig]
      {\goodCtx{\Gamma}{i} \quad n \in \nat}
      {\goodSig{\Gamma}{\_}{n} \text{ is well-formed}}
      $$
      $$
      \Rule[name=Seal]
      {\goodCtx{\Gamma}{i} \quad \goodSig{\Gamma}{\sigma}{n} 
      \quad \goodType{\Gamma}{A}{} }
      {\goodSeal{\Gamma}{\_}{\sigma}{A} \text{ is well-formed}}
      $$
  \end{minipage}
  \begin{minipage}[b]{0.6\linewidth}
    \begin{minted}[]{agda}
      data Con   : Set 
      data Ty    : Con → Set   
      data Tms   : Con → Con → Set 
      data Tm    : (Γ : Con) → Ty Γ  → Set 
      data Sig  : Con → ℕ → Set
      data WSig : Con → ℕ → Set 
      data Inh  : (Γ : Con) → Sig Γ n → Sig Γ m → Set
      Seal : (Γ : Con) → Sig Γ n → Sig Γ n → Set
    \end{minted}
  \end{minipage}
\begin{align*}
  &\Gamma~:~"Con"&&\mathboxtext{\goodCtx{\Gamma}{}} &&T~:~"Ty"~\Gamma&&\mathboxtext{\goodType{\Gamma}{T}{}}  &&\sigma~:~"Tms"~\Gamma~\Delta&&\mathboxtext{\goodSub{\Gamma}{\sigma}{\Delta}}  \\ & t~:~"Tm"~\Gamma~T&&\mathboxtext{\goodTerm{\Gamma}{t}{T}} 
  &&\sigma~:~"Sig"~\Gamma~n&&\mathboxtext{\goodSig{\Gamma}{\sigma}{n}}  &&\sigma~:~"WSig"~\Gamma~n&&\mathboxtext{\goodWSig{\Gamma}{\sigma}{n}} \\ &h~:~"Inh"~\Gamma~\sigma~\tau&&\mathboxtext{\goodInh{\Gamma}{h}{\sigma}{\tau}} && f~:~"Seal"~\Gamma~\sigma~A&&\mathboxtext{\goodSeal{\Gamma}{f}{\sigma}{A}}
\end{align*}
\caption{Well-formedness of Four New Judgements, and Agda-encoding-Judgements correspondence}
\end{figure}

@@

Now we show selected typing rules: we will have dependent function types
and dependent pair types. Dependent pair type is used to model Coq
module, which is useful when we need to model the ``compilation'' from
linkages to modules. We also have $\beta,\eta$-rule for both dependent
function type and dependent pair type.

@@

$$
\Rule[name=Ind Type]
{\goodTerm{\Gamma}{T}{\bW \sigma}}
{\goodTerm{\Gamma}{"W" \ T}{\cU}}
\quad
\Rule[name=Ind Term]
{\goodTerm{\Gamma}{T}{\bW \sigma}
  \quad \goodTerm{\Gamma}{a}{\pi^j_1\sigma}
  \quad \goodTerm{\Gamma, \pi^j_2\sigma[(id, a)]}{b}{El~("W"~T)}}
{\goodTerm{\Gamma}{"Wsup" \ T \ a \ b}{El~("W"~T)} }
$$

@@

In our meta-theory, we only formulate non-dependent eliminator, i.e. the
\textit{recursor}. Each recursor is constructed by a linkage containing
functions/handlers/pattern-matching cases dealing with each
constructor. Thus code reuse for recursor is delegated to linkage
inheritance---to reuse one particular pattern-matching case,
we just inherit the function/handler. This greatly simplifies the
meta\-theory development.

@@

As we said earlier, $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is actually just
a short hand of term judgement $\goodTerm{\Gamma , \cC
\sigma}{\_}{A[\pi_1]}$. The substitution law should be understood as a
lemma, and it is easily derivable using $\gamma^\uparrow$.

@@

With the definition of sealing judgement in mind, we can see
\ruleref{Sig Add} and \ruleref{Lnkg Add} work exactly as the
implementation. In the implementation, we will have "self__" all around
when defining new fields in our plugin, and this "self__" is exactly the
$A$ in the context of the two fields -- especially when we don't do any
sealing, this $A = \cC \sigma$, which is exactly how the implementation
exposes "self__" as a module parameter.
